---
title: "Previous Findings"
subtitle: "Convergence between locus|gene pairs reported on this analysis and previously reported colocalizations (BigBrain, Bellenguez and SingleBrain papers)."
format:
  html:
    page-layout: full
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
pkgs <- c("data.table", "stringr", "DT", "htmltools", "eulerr")
to_install <- setdiff(pkgs, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install, repos = "https://cloud.r-project.org")
library(data.table); library(stringr); library(DT); library(htmltools); library(eulerr)

# arquivo bruto: mesma tabela "lista_locus" já usada antes pra calcular
# convergência -- ver data/README dentro deste site
L <- fread("data/lista_locus.csv")
L <- L[!is.na(locus)]

split_genes <- function(x) {
  if (is.na(x) || x == "") return(character(0))
  p <- str_trim(str_split(x, "[,;]")[[1]])
  p[p != "" & !toupper(p) %in% c("NA", "NAN", "-")]
}

lit_cols <- intersect(c("bigbrain_eqtl", "bigbrain_sqtl", "genes_bellenguez", "singlebrain_results"), names(L))

L[, meus := lapply(genes_significant, split_genes)]
L[, lit  := lapply(seq_len(.N), function(i) {
  Reduce(union, lapply(lit_cols, function(cc) split_genes(L[[cc]][i])))
})]
L[, inter := Map(intersect, meus, lit)]
L[, novos := Map(setdiff, meus, lit)]

L[, has_mine := lengths(meus) > 0]
L[, has_lit  := lengths(lit)  > 0]

# Quatro categorias mutuamente exclusivas -- cada locus cai em exatamente uma
# região do diagrama de Euler abaixo.
L[, status := fcase(
  !has_lit,               "No comparison",
  lengths(inter) > 0,     "Concordant",
  has_mine,               "Discordant",
  default = "Literature only"
)]

kpi <- function(num, lab) div(class = "kpi-box", span(class = "num", num), span(class = "lab", lab))
comparaveis <- L[status != "No comparison"]
```

```{r kpis}
div(
  kpi(nrow(L), "Loci on the table"),
  kpi(nrow(comparaveis), "with literature reports to compare"),
  kpi(paste0(round(mean(comparaveis$status == "Concordant") * 100), "%"), "concordance"),
  kpi(sum(L$status == "No comparison"), "no previous register (new or non-audited loci)")
)
```

## Overview

```{r euler}
#| fig-width: 8
#| fig-height: 4.5
#| fig-align: center

regions <- c(
  "This study"                 = sum(L$status == "Discordant"),
  "Literature"                 = sum(L$status == "Literature only"),
  "This study&Literature"      = sum(L$status == "Concordant"),
  "New (no literature)"        = sum(L$status == "No comparison")
)
regions <- regions[regions > 0]

# euler() lê um vetor nomeado como contagens de regiões DISJUNTAS: "This study"
# é a área exclusiva, "This study&Literature" é apenas a interseção. O terceiro
# conjunto não tem interseção declarada, então é desenhado como bolha isolada.
fit <- eulerr::euler(regions)

# uma cor por conjunto desenhado -- se alguma categoria zerar, o conjunto some
# do fit e passar 3 cores fixas quebraria o plot
fill_pal <- c("#1F6FB4", "#E4A33C", "#7FA96B")[seq_len(nrow(fit$ellipses))]

# Os rótulos nas próprias formas ficam cortados: a região exclusiva de "This
# study" é um crescente colado na borda esquerda, então o texto sai do device.
# Legenda lateral resolve sem precisar mexer no ajuste do Euler.
plot(fit,
     quantities = list(fontsize = 15, col = "grey15"),
     labels     = FALSE,
     legend     = list(side = "right", fontsize = 12),
     fills      = list(fill = fill_pal, alpha = 0.55),
     edges      = list(col = "grey30", lwd = 1.2))
```

*Overlap = loci where at least one gene nominated here was also reported in the literature (concordant). "This study" alone = loci reported in the literature, but with no gene in common (discordant). "Literature" alone = loci with a previous report that did not reach significance here. The isolated bubble holds loci with no literature counterpart at all.*

## Full table

```{r tabela}
tab <- L[, .(
  locus, cohort = descoberta_loci, status,
  results = vapply(meus, paste, character(1), collapse = ", "),
  literature  = vapply(lit, paste, character(1), collapse = ", "),
  coincide   = vapply(inter, paste, character(1), collapse = ", "),
  additional  = vapply(novos, paste, character(1), collapse = ", ")
)]
setorder(tab, status, locus)

datatable(tab, rownames = FALSE, filter = "top", extensions = "Buttons",
          options = list(pageLength = 20, dom = "Bfrtip", buttons = c("copy", "csv"),
                         scrollX = TRUE),
          class = "compact stripe")
```

::: {.callout-note collapse="true"}
## How concordance is calculated

`Concordant` = at least one gene in common between the genes that colocalized on this analysis (`genes_significant`) and the union of literature sources (BigBrain eQTL/sQTL, Bellenguez, SingleBrain). `Discordant` = there's literature for the locus, but none of the genes coincides. `Literature only` = the locus was reported before, but nothing reached significance here. `No comparison` = none of the sources covers this locus (includes all 19 new loci from consensus meta-analysis).
:::
