---
title: "Previous Findings"
subtitle: "Convergence between locus|gene pairs reported on this analysis and previously reported colocalizations (BigBrain, Bellenguez and SingleBrain papers)."
format:
  html:
    page-layout: full
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(data.table); library(stringr); library(DT); library(htmltools)

# arquivo bruto: mesma tabela "lista_locus" já usada antes pra calcular
# convergência -- ver data/README dentro deste site
L <- fread("data/lista_locus.csv")
L <- L[!is.na(locus)]

split_genes <- function(x) {
  if (is.na(x) || x == "") return(character(0))
  p <- str_trim(str_split(x, "[,;]")[[1]])
  p[p != "" & !toupper(p) %in% c("NA", "NAN", "-")]
}

lit_cols <- intersect(c("bigbrain_eqtl", "bigbrain_sqtl", "genes_bellenguez", "singlebrain_results"), names(L))

L[, meus := lapply(genes_significant, split_genes)]
L[, lit  := lapply(seq_len(.N), function(i) {
  Reduce(union, lapply(lit_cols, function(cc) split_genes(L[[cc]][i])))
})]
L[, inter := Map(intersect, meus, lit)]
L[, novos := Map(setdiff, meus, lit)]

L[, status := fcase(
  lengths(lit) == 0, "No comparison",
  lengths(inter) > 0, "Convergence",
  default = "Divergence"
)]

kpi <- function(num, lab) div(class = "kpi-box", span(class = "num", num), span(class = "lab", lab))
comparaveis <- L[status != "No comparison"]
```

```{r kpis}
div(
  kpi(nrow(L), "Loci on the table"),
  kpi(nrow(comparaveis), "with literature reports to compare"),
  kpi(paste0(round(mean(comparaveis$status == "Convergence") * 100), "%"), "convergence"),
  kpi(sum(L$status == "No comparison"), "no previous register (new or non-audited loci)")
)
```

## Full table

```{r tabela}
tab <- L[, .(
  locus, cohort = descoberta_loci, status,
  n_significant, n_ambiguous,
  results = vapply(meus, paste, character(1), collapse = ", "),
  literature  = vapply(lit, paste, character(1), collapse = ", "),
  coincide   = vapply(inter, paste, character(1), collapse = ", "),
  additional  = vapply(novos, paste, character(1), collapse = ", ")
)]
setorder(tab, status, locus)

datatable(tab, rownames = FALSE, filter = "top", extensions = "Buttons",
          options = list(pageLength = 20, dom = "Bfrtip", buttons = c("copy", "csv"),
                         scrollX = TRUE),
          class = "compact stripe")
```

::: {.callout-note collapse="true"}
## Como a convergência é calculada

`Convergence` = at least one gene in common between the genes that colocalized on this analysis (`genes_significant`) and the union of literature sources (BigBrain eQTL/sQTL, Bellenguez, SingleBrain). `Divergence` = there's literature for the locus, but none of the genes coincides. `No comparison` = none of the sources covers this locus (includes all 19 new loci from consensus meta-analysis).
:::
